\chapter{Acknowledgements}

Thank you to everyone who has supported me during throughout my PhD. 

\vspace{5mm}

This research would not have been possible without the support and guidance from my supervisors, David MacTaggart and Radostin Simitev. David, thank you for your patience and continual feedback. You've given an immense amount of time and energy to guiding my research, you've provided focus and insight and you've nurtured in me a thoroughness I did not previously possess. Rado, you've been a mentor, a stoic mediator and an inspiration. If I end up with a career half as illustrious as yours I will be happy.

Thank you also to the staff of the school of Maths \& Stats, in particular Dave for sorting out many a digital niggle; Jean, Margaret, Pauline, and Sharon for the care they take in dealing with every minutiae of admin; and Steve and Rob for convincing me to stick through the tough periods.

I would also like to thank the EPSRC for supporting me financially through my PhD, and the group behind ARCHIE-WeSt at the University of Strathclyde for their work in supporting the computational needs of Glasgow's academia.

\vspace{5mm}

I would be much closer to mental ruin without the support of my parents, Maureen and Patrick, and my partner in crime, Annie. Mum and Dad, thank you for encouraging me to choose my own path and pretending to know what I'm talking about when I warble on about the solar atmosphere. Annie, you make me food and make me laugh. Thank you for being who you are.

\vspace{5mm}

To my fellow PhD students in the School of Maths \& Stats, particularly Kellan, Flynn, Pete, Roxy, Jay, Luke and Anna, thank you for holidays in Lanzarote and Frankfurt, for silly arguments and hilarious stories.
