\chapter{The effect of viscosity on the Kelvin-Helmholtz instability in the fan plane of a null point}

\section{Results}

\subsection{Parameter study}

A parameter study was run, varying $\nu$ over three values ($10^{-5}$, $10^{-4}$ and $10^{-3}$), and varying $\eta$ over the same three values, giving nine parameter sets in total. For each of these, two simulations were run, one using the isotropic model and the other using the switching model.

**General results**

There are some features common to all simulations: the initial interaction of the twisting motion, resultant counterflows, and the eventual collapse of the null. Initially we observe torsional Alfv\'en waves propagating from the upper and lower footpoints along the field and into the fan plane. The counterflows present in~\cite{wyperKelvinHelmholtzInstabilityCurrentvortex2013} are also seen at $t=3$ TODO PLOT. Much later in the evolution (depending on the value of $\eta$) the null begins to collapse during a period of intense reconnection PLOTS TODO.

The Kelvin-Helmholtz instability only occurs in those simulations with suitably low viscosity, and a suitable (TODO what is suitable?) early-time counterflow. These conditions are met in all switching cases, except when $\eta=10^{-5}$, and only in the isotropic case of $\nu = 10^{-5}, \eta=10^{-3}$. In all KH unstable cases, the initial instability is seen early, around $t=1$ PLOT TODO, and peaks around $t=10$ TODO CHECK THIS. The evolution takes different forms for different values of $\eta$ and $\nu$.

- The nature of the reconnection is violent and produces a great deal of Ohmic heating
- The time at which the reconnection occurs isn't hugely affected by the form and strength of viscosity. The major factor affecting the onset time is the resistivity; the lower the value, the sooner the onset.
- The resistivity seems to control when reconnection starts; the lower the resistivity the sooner the reconnection begins
- For low ($\eta=1e-5$) resistivity the KHI in the usually unstable switching case is disrupted. Interestingly, it's not the eventual reconnection that disrupts, but the instability isn't even able to start. It's not clear why.
